Approximation as Differentiation

### 2025-07-10 main/language.tex
Rough pass.

A model $M$ of $\Sigma$ can be given in any category $\cat{C}$ with finite products and a terminal object, and
assigns to each base type $\rho \in \PrimTy$ an object $\sem{\rho}_{M}$ in $\cat{C}$, and to each primitive
operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}_{M}: \sem{\rho_1}_{M} \times
\ldots \times \sem{\rho_n}_{M} \to \sem{\rho}_{M}$. A model of $\tyList$ can also be given in such a category,
and consists of an endofunctor $\List: \cat{C} \to \cat{C}$, plus for any object $X$, morphisms $\nil: 1 \to
\List(X)$ and $\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms $f_\nil: Z \to
Y$ and $f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times \List(X) \to Y$,
satisfying the usual laws for lists.

To give the semantics for the full language, we fix a bicartesian closed category $\Sem$ with a strong monad
$(\Lift, \eta, \mu)$ with strength $\mathsf{st}$, a model $M$ of $\Sigma$ in $\Sem$, and a model of $\tyList$
in $\Sem$. \figref{semantics:types} gives the interpretations of types $\tau$ and contexts $\Gamma$ as objects
of $\Sem$. \figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as morphisms
$\sem{\Gamma} \to \sem{\tau}$.

@@

Given a model $M$ of $\Sigma$ in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times \JoinSLat^\op)$ can
implement $\Sem$. It is bicartesian closed by \corref{mslat-jslat-bcc}, and can model $\List(X)$ as an
infinite coproduct (\secref{models-of-total-gps:coproducts-and-products}). A model $H(M)$ of $\Sigma$ can be
obtained via the embedding $\HoEmbed: \Fam(\LatGal) \to \Fam(\MeetSLat \times \JoinSLat^\op)$ given in
\propref{ho-embedding}, because .... Finally $\LatGal$ has a lifting monad which is preserved by the 2-functor
$\Fam: \Cat \to \Cat$ (\secref{first-order:lifting}). The correctness of this interpretation is considered in
the next section.

>>>>

A model $M$ of $\Sigma$ can be given in any category $\cat{C}$ with finite products, and assigns to each base
type $\rho \in \PrimTy$ an object $\sem{\rho}_{\PrimOp}$ in $\cat{C}$, and to each primitive operation $\phi
\in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}_{\Op}: \sem{\rho_1}_{\PrimTy} \times \ldots
\times \sem{\rho_n}_{\PrimTy} \to \sem{\rho}_{\PrimTy}$. A model of $\tyList$ can also be given in such a
category, and consists of an endofunctor $\List: \cat{C} \to \cat{C}$, plus for any object $X$, morphisms
$\nil: 1 \to \List(X)$ and $\cons: X \times \List(X) \to \List(X)$, and for any objects $Y, Z$ and morphisms
$f_\nil: Z \to Y$ and $f_\cons: Z \times X \times Y \to Y$, a morphism $\fold(f_\nil,f_\cons): Z \times
\List(X) \to Y$, with $\nil$, $\cons$ and $\fold$ satisfying the usual laws for lists.

The semantics for the full language is then given in terms of an arbitrary bicartesian closed category $\Sem$,
models of $\Sigma$ and $\tyList$ in $\Sem$, and a strong monad $(\Lift, \eta, \mu)$ with strength
$\mathsf{st}$ for interpreting $\tyLift$. \figref{semantics:types} gives the interpretations of types $\tau$
and contexts $\Gamma$ as objects; \figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t:
\tau$ as morphisms $\sem{\Gamma} \to \sem{\tau}$.

@@

Given a model $M$ of $\Sigma$ in $\Fam(\LatGal)$, the category $\Fam(\MeetSLat \times \JoinSLat^\op)$ can
implement $\Sem$. It is bicartesian closed by \corref{mslat-jslat-bcc}, and can model $\tyList$ using infinite
coproducts (\secref{models-of-total-gps:coproducts-and-products}). The embedding $\HoEmbed: \Fam(\LatGal) \to
\Fam(\MeetSLat \times \JoinSLat^\op)$ from \propref{ho-embedding} preserves finite products, providing a model
$H(M)$ of $\Sigma$. Finally the lifting monad in $\Fam(\LatGal)$ (\secref{first-order:lifting}) is also
preserved by $H$. The correctness of this interpretation is considered in the next section.
